---
title: "Darwin Core mapping"
subtitle: "For: Bryophyte checklist and European red list of the Brussels-Capital Region, Flanders and Wallonia (Belgium)"
author:
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/rinse-pathways-checklist/blob/master/src/dwc_mapping.Rmd).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(digest)    # To generate hashes
library(rgbif)     # To parse names
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/bryophytes.xlsx"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
dwc_description_file = "../data/processed/description.csv"
```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
raw_data <- read_excel(path = raw_data_file) 
```

Clean the data somewhat: remove empty rows if present and clean names:

```{r}
raw_data %<>% 
  remove_empty("rows") %<>%
  clean_names()
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable. S

Generate `taxonID`:

```{r}
# Vectorize the digest function (the digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxonID = paste("bryophytes-belgium-checklist", "taxon", vdigest (taxon, algo = "md5"), sep = ":"))
```

## Further pre-processing

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/") 
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "Botanical Garden Meise")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### institutionCode

```{r}
taxon %<>% mutate(institutionCode = "BGM") 
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Bryophyte checklist and European red list of the Brussels-Capital Region, Flanders and Wallonia (Belgium)") 
```

### taxonID	

```{r}
taxon %<>% mutate(taxonID = raw_taxonID) 
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = "Plantae") 
```

### phylum

```{r}
taxon %<>% mutate(phylum = raw_division) 
```

### taxonRank

Almost all taxa are species, only `Tortella inclinata (R.Hedw.) Limpr. var. densa` is a variety:


```{r}

```

### nomenclaturalCode

